Some examples in extended logic counting classes

### 2017-01-05 LICS/counting.tex
sin comentarios
04 ene

Fix a relational signature $\R = \{R_1, \ldots, R_k\}$ and assume that $\A$ is an ordered finite $\R$-structure with a domain $A$ containing $n$ elements. Recall that  $<$ is a linear order on $A$, say $a_1 < a_2 < \ldots < a_n$. For every $i \in \{1, \ldots, k\}$, define the encoding of $R_i^\A$, denoted by $\enc(R_i^\A)$, as the following binary string. Assume that $\ell = \arity(R_i)$ and consider an enumeration of the $\ell$-tuples over $A$ in the lexicographic order induced by $<$. 
%(that is, $(a_1, \ldots, a_1, a_1)$, $(a_1, \ldots, a_1, a_2)$, $\ldots$, $(a_n, \ldots, a_n, a_{n-1})$, $(a_n, \ldots, a_n, a_n)$). 
Then let $\enc(R_i^\A)$ be a binary string of length $n^\ell$ such that the $i$-th bit of $\enc(R_i^\A)$ is 1 if the $i$-th tuple in the previous enumeration belongs to $R_i^\A$, and 0 otherwise. Moreover, define the encoding of $\A$, denoted by $\enc(\A)$, as the following binary string~\cite{L04}:
\begin{eqnarray*}
	\enc(\A) & = & 0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A).
\end{eqnarray*}
%We define the class of all $\R$-functions, denoted by $\Func(\R)$, as the class of all functions $f: \ostr \rightarrow \bbN$.
%Given a function complexity class $\CC$ (i.e. $f: \Sigma^* \rightarrow \bbN$ for every $f \in \CC$), we say that a function $f \in \Func(\R)$ can be computed in $\CC$ if there exists $g \in \CC$ such that $f(\A) = g(\enc(\A))$ for every $\A \in \ostr$. 
%Note that the function $g$ outputs $f$ for encodings of structures and can behave arbitrarily otherwise.
We can now formalize the notion of capturing a function complexity class over $\R$-structures.
\begin{definition} \label{cap-def}
	Let $\FF$ be a fragment of $\qso$ and $\CC$ a function complexity class. Then we say that {\em  $\FF$ captures $\CC$ over $\R$-structures} if the  following conditions are true:
	\begin{enumerate}
		\item for every $\alpha \in \FF$, there exists $f \in \CC$ such that $\sem{\alpha}(\A) = f(\enc(\A))$ for every $\A \in \ostr[\R]$.

		\item for every $f \in \CC$, there exists $\alpha \in \FF$ such that   $f(\enc(\A)) = \sem{\alpha}(\A)$ for every $\A \in \ostr[\R]$.
	\end{enumerate} 
	Finally, we say that $\FF$ captures $\CC$ if $\FF$ captures $\CC$ over $\R$-structures for every relational signature $\R$.
\end{definition}
%For the sake of simplification, we denote the first condition by $\FF \subseteq \CC$ and the second condition by $\CC \subseteq \FF$.
Note that in Definition~\ref{cap-def} the function $f \in \CC$ and the formula $\alpha \in \FF$ must coincide in all the strings that encode $\R$-structures. Since we want to capture $\CC$ over a fix set of structures (e.g. graphs, matrices), it is natural to just consider strings that encodes $\R$-structures. 
This definition is fairly standard in descriptive complexity \cite{L04,immerman2012descriptive} and all notions for capturing complexity classes restrict $f \in \CC$ similarly.

@@

Instead, if we move from $\eqso$ we have other options for counting like first- and second-order products. Actually, if we combine $\qfo$ with $\lfp$ this is exactly what we need to capture $\fp$, the class of function that can be computed in~$\ptime$.
\begin{theorem} \label{theo:capture-fp}
	$\qfo(\lfp)$ captures $\fp$ over ordered structures.
\end{theorem}
To capture $\fp$ one first shows that any formula in $\qfo(\lfp)$ can be evaluated in $\ptime$. 
Indeed, $\lfp$ is a polynomial time logic~\cite{Immerman82,vardi1982complexity} and the sum and product quantifiers can be computed in $\fp$. 
For the other direction, one has to use $\lfp$ to simulate the run of the $\fp$-machine and use the quantitative quantifiers to reconstruct the binary number built in the tape. 
It is important to notice that the alternation between sum and products quantifiers is crucial for capturing $\fp$.
That is, without these operators is difficult to capture $\fp$.

>>>>

Fix a relational signature $\R = \{R_1, \ldots, R_k\}$ and assume that $\A$ is an ordered finite $\R$-structure with a domain $A$ containing $n$ elements. Recall that  $<$ is a linear order on $A$, say $a_1 < a_2 < \ldots < a_n$. For every $i \in \{1, \ldots, k\}$, define the encoding of $R_i^\A$, denoted by $\enc(R_i^\A)$, as the following binary string. Assume that $\ell = \arity(R_i)$ and consider an enumeration of the $\ell$-tuples over $A$ in the lexicographic order induced by $<$. 
%(that is, $(a_1, \ldots, a_1, a_1)$, $(a_1, \ldots, a_1, a_2)$, $\ldots$, $(a_n, \ldots, a_n, a_{n-1})$, $(a_n, \ldots, a_n, a_n)$). 
Then let $\enc(R_i^\A)$ be a binary string of length $n^\ell$ such that the $i$-th bit of $\enc(R_i^\A)$ is 1 if the $i$-th tuple in the previous enumeration belongs to $R_i^\A$, and 0 otherwise. Moreover, define the encoding of $\A$, denoted by $\enc(\A)$, as the following binary string~\cite{L04}:
\begin{eqnarray*}
	\enc(\A) & = & 0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A).
\end{eqnarray*}
%We define the class of all $\R$-functions, denoted by $\Func(\R)$, as the class of all functions $f: \ostr \rightarrow \bbN$.
%Given a function complexity class $\CC$ (i.e. $f: \Sigma^* \rightarrow \bbN$ for every $f \in \CC$), we say that a function $f \in \Func(\R)$ can be computed in $\CC$ if there exists $g \in \CC$ such that $f(\A) = g(\enc(\A))$ for every $\A \in \ostr$. 
%Note that the function $g$ outputs $f$ for encodings of structures and can behave arbitrarily otherwise.
We can now formalize the notion of capturing a function complexity class over $\R$-structures.
\begin{definition} \label{def:cap}
	Let $\FF$ be a fragment of $\qso$ and $\CC$ a function complexity class. Then we say that {\em  $\FF$ captures $\CC$ over $\R$-structures} if the  following conditions are true:
	\begin{enumerate}
		\item for every $\alpha \in \FF$, there exists $f \in \CC$ such that $\sem{\alpha}(\A) = f(\enc(\A))$ for every $\A \in \ostr[\R]$.

		\item for every $f \in \CC$, there exists $\alpha \in \FF$ such that   $f(\enc(\A)) = \sem{\alpha}(\A)$ for every $\A \in \ostr[\R]$.
	\end{enumerate} 
	Finally, we say that $\FF$ captures $\CC$ if $\FF$ captures $\CC$ over $\R$-structures for every relational signature $\R$.
\end{definition}
%For the sake of simplification, we denote the first condition by $\FF \subseteq \CC$ and the second condition by $\CC \subseteq \FF$.
Note that in Definition~\ref{def:cap} the function $f \in \CC$ and the formula $\alpha \in \FF$ must coincide in all the strings that encode $\R$-structures. Since we want to capture $\CC$ over a fix set of structures (e.g. graphs, matrices), it is natural to just consider strings that encodes $\R$-structures. 
This definition is fairly standard in descriptive complexity \cite{L04,immerman2012descriptive} and all notions for capturing complexity classes restrict $f \in \CC$ similarly.

@@

Instead, if we move from $\eqso$ we have other options for counting like first- and second-order products. Actually, if we combine $\qfo$ with $\lfp$ this is exactly what we need to capture $\fp$, the class of function that can be computed in~$\ptime$.
\begin{theorem} \label{theo:capture-fp}
	$\qfo(\lfp)$ captures $\fp$ over ordered structures.
\end{theorem}
To capture $\fp$ one first shows that any formula in $\qfo(\lfp)$ can be evaluated in $\ptime$. 
Indeed, $\lfp$ is a polynomial time logic~\cite{I86,vardi1982complexity} and the sum and product quantifiers can be computed in $\fp$. 
For the other direction, one has to use $\lfp$ to simulate the run of the $\fp$-machine and use the quantitative quantifiers to reconstruct the binary number built in the tape. 
It is important to notice that the alternation between sum and products quantifiers is crucial for capturing $\fp$.
That is, without these operators is difficult to capture $\fp$.

### 2017-01-05 LICS/syntactic-clo.tex
04 ene

It was proved in \cite{SalujaST95} that every function in $\E{1}$ admits an FPRAS. Besides, it can be proved that $\E{1} \subseteq \totp$. However, this class is not closed under sum as shown in the following proposition.
\begin{proposition}\label{prop-e1-nc}
There exist functions $f, g \in \E{1}$ such that $(f + g) \not\in \E{1}$.
\end{proposition}
To overcome this limitation, one can consider the class $\QE{1}$, which is closed by under sum by definition. In fact, the following proposition shows that the same good properties as for $\E{1}$ hold for $\QE{1}$.
\begin{proposition} \label{prop:qe0-fp-qe1-totp-fptras}
$\QE{1} \subseteq \totp$ and every function in $\QE{1}$ has an FPRAS.
\end{proposition}
Moreover, from the following theorem one can conclude that $\QE{1}$ is closed under multiplication.
\begin{theorem}\label{theo:binary-prod}
	If $\LL$ is closed under conjunction, then $\eqso(\LL)$ is closed under multiplication.
\end{theorem}
Hence, it only remains to prove that $\QE{1}$ is closed under subtraction by one. Unfortunately, it is not clear whether this property holds; in fact, we conjecture that it is not the case. Thus, we need to find an extension of $\QE{1}$ that keeps all the previous properties and is closed under subtraction by one. It is important to notice that $\shp$ is believed not to be closed under subtraction by one by some complexity-theoretical assumption,\footnote{A decision problem $L$ is in the randomized complexity class $\cspp$ if there exists a nondeterministic polynomial-time Turing Machine $M$ such that for every $x \in L$ it holds that $\tma_M(x) - \tmr_M(x) = 2$, and for every $x \not\in L$ it holds that $\tma_M(x) = \tmr_M(x)$ \cite{OH93,FFK94}. It is believed that $\np \not\subseteq \cspp$, as there is no evidence that a Turing Machine with the property just described exists for the propositional satisfiability problem. However, if $\shp$ is closed under subtraction by one, then it holds that $\np \subseteq \cspp$ \cite{OH93}.} so the following proposition rules out any logic that extends $\Pi_1$.
\begin{proposition} \label{pi-minusone}
If $\Pi_1 \subseteq \LL$ and $\eqso(\LL)$ is closed under subtraction by one, then $\shp$ is closed under subtraction by one. 
\end{proposition}
\martin{para este teorema se necesita que $\LL$ sea un fragmento de $\fo$}
Therefore, the desired extension has to be achieved by allowing some local extensions to $\Sigma_1$. More precisely, we define $\logex{1}$ as $\Sigma_1$ but allowing atomic formulas over a relational signature $\R$ to be of the form either $u = v$ or $X(\bar u)$, where $X$ is a second-order variable, or $\varphi(\bar u)$, where $\varphi(\bar u)$ is a first-order formula over $\R$ (in particular, it does not mentioned any second-order variable). With this extension we have that:
\begin{theorem}\label{sigmafo-minusone}
The class $\eqso(\logex{1})$ is closed under sum, product and subtraction by one. Moreover, $\eqso(\logex{1}) \subseteq \totp$ and every function in $\eqso(\logex{1})$ has an FPRAS.
\end{theorem}

>>>>

It was proved in \cite{SalujaST95} that every function in $\E{1}$ admits an FPRAS. Besides, it can be proved that $\E{1} \subseteq \totp$. However, this class is not closed under sum as shown in the following proposition.
\begin{proposition}\label{prop-e1-nc}
There exist functions $f, g \in \E{1}$ such that $(f + g) \not\in \E{1}$.
\end{proposition}
To overcome this limitation, one can consider the class $\QE{1}$, which is closed by under sum by definition. In fact, the following proposition shows that the same good properties as for $\E{1}$ hold for $\QE{1}$.
\begin{proposition} \label{prop:qe0-fp-qe1-totp-fptras}
$\QE{1} \subseteq \totp$ and every function in $\QE{1}$ has an FPRAS.
\end{proposition}
Moreover, from the following theorem one can conclude that $\QE{1}$ is closed under multiplication.
\begin{theorem}\label{theo:binary-prod}
	If $\LL$ is closed under conjunction, then $\eqso(\LL)$ is closed under multiplication.
\end{theorem}
Hence, it only remains to prove that $\QE{1}$ is closed under subtraction by one. Unfortunately, it is not clear whether this property holds; in fact, we conjecture that it is not the case. Thus, we need to find an extension of $\QE{1}$ that keeps all the previous properties and is closed under subtraction by one. It is important to notice that $\shp$ is believed not to be closed under subtraction by one by some complexity-theoretical assumption,\footnote{A decision problem $L$ is in the randomized complexity class $\cspp$ if there exists a nondeterministic polynomial-time Turing Machine $M$ such that for every $x \in L$ it holds that $\tma_M(x) - \tmr_M(x) = 2$, and for every $x \not\in L$ it holds that $\tma_M(x) = \tmr_M(x)$ \cite{OH93,FFK94}. It is believed that $\np \not\subseteq \cspp$, as there is no evidence that a Turing Machine with the property just described exists for the propositional satisfiability problem. However, if $\shp$ is closed under subtraction by one, then it holds that $\np \subseteq \cspp$ \cite{OH93}.} so the following proposition rules out any logic that extends $\Pi_1$.
\begin{proposition} \label{pi-minusone}
If $\Pi_1 \subseteq \LL$ and $\eqso(\LL)$ is closed under subtraction by one, then $\shp$ is closed under subtraction by one. 
\end{proposition}
\martin{para este resultado se necesita que $\LL$ sea un fragmento de $\fo$}
Therefore, the desired extension has to be achieved by allowing some local extensions to $\Sigma_1$. More precisely, we define $\logex{1}$ as $\Sigma_1$ but allowing atomic formulas over a relational signature $\R$ to be of the form either $u = v$ or $X(\bar u)$, where $X$ is a second-order variable, or $\varphi(\bar u)$, where $\varphi(\bar u)$ is a first-order formula over $\R$ (in particular, it does not mentioned any second-order variable). With this extension we have that:
\begin{theorem}\label{sigmafo-minusone}
The class $\eqso(\logex{1})$ is closed under sum, product and subtraction by one. Moreover, $\eqso(\logex{1}) \subseteq \totp$ and every function in $\eqso(\logex{1})$ has an FPRAS.
\end{theorem}
